Practical Optional Types for Clojure

### 2015-10-16 formal-model.tex
shaved 1/4 page

Now that we have demonstrated the core features Typed Clojure
provides, we link them together in a formal model called
\lambdatc{}.
%
Building on occurrence typing,
we incrementally add each
novel feature of Typed Clojure to the formalism,
interleaving presentation of syntax, typing rules, operational semantics,
and subtyping.

@@

Our presentation will start with a review of
occurrence typing~\cite{TF10}, the foundation of \lambdatc{}.
%We build up the occurrence typing calculus for illustrative purposes, 
%and present the full syntax at the end of the section.

@@

To illustrate, recall \egref{example:desserts-on-meal}.
We know the parameter \makelocal{o} is of type $\Order$,
%by the annotation on $desserts$
written
{\isprop{\Order}{\makelocal{o}}}
as a proposition.
%
In checking the ${\makekw{combo}}$ method, we also assume
${\appexp{\makekw{Meal}}{\makelocal{o}}}$ is ${\makekw{combo}}$,
based on multimethod dispatch rules. This is written
  {\isprop{\Value{\makekw{combo}}}{\path{\keype{\makekw{Meal}}}{\makelocal{o}}}},
pronounced ``the ${\makekw{Meal}}$ path of variable \makelocal{o} is of type
{\Value{\makekw{combo}}}''.

@@

We formalize our system following Tobin-Hochstadt and Felleisen \cite{TF10}.
%(with differences highlighted in $\trdiff{\text{blue}}$)
The typing judgment 
$$
{\judgementrewrite   {\propenv}
              {\e{}} {\t{}}
  {\filterset {\thenprop {\prop{}}}
              {\elseprop {\prop{}}}}
  {\object{}}
  {\ep{}}}
$$
says expression \e{} rewrites to \ep{}, which
is of type \t{} in the 
proposition environment $\propenv{}$, with 
`then' proposition {\thenprop {\prop{}}}, `else' proposition {\elseprop {\prop{}}}
and object \object{}.

@@

The lambda rule T-Abs introduces \isprop{\s{}}{\x{}}} = \isprop{\Top}{\makelocal{d}}
to check the body.
With \propenv{} = \isprop{\Top}{\makelocal{d}},
the T-If rule first checks the test \e{1} = {\appexp{\numberhuh{}}{\makelocal{d}}}
via the T-App rule, with three main steps.

@@

Resuming {\appexp{\numberhuh{}}{\makelocal{d}}},
in T-App the operand \ep{} = \makelocal{d} is checked with
T-Local as
\begin{equation}
\judgementselfrewrite{\propenv{}}
                     {\makelocal{d}}
                     {\Top}
                     {\filterset{\notprop{\falsy}{\makelocal{d}}}
                                {\isprop{\falsy}{\makelocal{d}}}}
                     {\makelocal{d}}
\end{equation}
which encodes the type, proposition, and object information
about variables
as previously discussed.

@@

We now walk through the conversion from unresolved expression~\ref{eq:unresolved} to 
resolved expression~\ref{eq:resolved}.

@@

\begin{figure}
  $$
\LUpdate{}
$$
\caption{Update rule}
\label{main:figure:updaterule}
\end{figure}

\begin{figure}
$$
\begin{array}{lr}
  \begin{array}{llll}
    \restrictfigure{}
  \end{array}
  \ \ 
  \begin{array}{llll}
    \removefigure{}
  \end{array}
\end{array}
$$
\caption{Restrict and remove}
\label{main:figure:restrictremove}
\end{figure}

@@

A lookup of a key that is neither present or absent is checked by
T-GetHMapPartialDefault.
$$
\abs{\makelocal{u}}{\Lunch}{\getexp{\makelocal{u}}{\makekw{bocce}}}
$$
The result type is \Top, and the return object is \path{\keype{\makekw{bocce}}}{\makelocal{u}}.
Notice propositions are erased once they enter a HMap type.

@@

Extending a map with T-AssocHMap preserves its completeness.
$$
\abs{\makelocal{b}}{\Breakfast}{\assocexp{\makelocal{b}}{\makekw{au}}{\makestr{beans}}}
$$
The result type is
$
\HMapc{\mandatoryset{\mandatoryentrynoarrow{\Valkw{en}}{\String}, {\mandatoryentrynoarrow{\Valkw{fr}}{\String}}
        ,{\mandatoryentrynoarrow{\Valkw{au}}{\String}}}}
$,
which is complete, like its input.
T-AssocHMap also enforces ${\k{}} \not\in {\absent{}}$ to prevent badly formed types.

@@

\paragraph{Subtyping}
Subtyping for HMaps
designate \MapLiteral{} as a common supertype for all HMaps.
S-HMap says that an HMap is a subtype of another HMap if they agree
on \completenessmeta{}, agree on mandatory entries with subtyping
and at least cover the absent keys of the supertype.
Complete maps are subtypes of partial maps
as long as they agree on the mandatory entries of the partial map via subtyping (S-HMapP).

>>>>

After demonstrating the core features Typed Clojure
provides, we link them together in a formal model called
\lambdatc{}.
%
Building on occurrence typing,
we incrementally add each
novel feature of Typed Clojure to the formalism,
interleaving presentation of syntax, typing rules, operational semantics,
and subtyping.

@@

We start with a review of
occurrence typing~\cite{TF10}, the foundation of \lambdatc{}.
%We build up the occurrence typing calculus for illustrative purposes, 
%and present the full syntax at the end of the section.

@@

To illustrate, recall \egref{example:desserts-on-meal}.
The parameter \makelocal{o} is of type $\Order$,
%by the annotation on $desserts$
written
{\isprop{\Order}{\makelocal{o}}}
as a proposition.
%
In the ${\makekw{combo}}$ method, we know
${\appexp{\makekw{Meal}}{\makelocal{o}}}$ is ${\makekw{combo}}$,
based on multimethod dispatch rules. This is written
  {\isprop{\Value{\makekw{combo}}}{\path{\keype{\makekw{Meal}}}{\makelocal{o}}}},
pronounced ``the ${\makekw{Meal}}$ path of variable \makelocal{o} is of type
{\Value{\makekw{combo}}}''.

@@

We formalize our system following Tobin-Hochstadt and Felleisen \cite{TF10}.
%(with differences highlighted in $\trdiff{\text{blue}}$)
The typing judgment 
$
{\judgementrewrite   {\propenv}
              {\e{}} {\t{}}
  {\filterset {\thenprop {\prop{}}}
              {\elseprop {\prop{}}}}
  {\object{}}
  {\ep{}}}
$
says expression \e{} rewrites to \ep{}, which
is of type \t{} in the 
proposition environment $\propenv{}$, with 
`then' proposition {\thenprop {\prop{}}}, `else' proposition {\elseprop {\prop{}}}
and object \object{}.

@@

The lambda rule T-Abs introduces \isprop{\s{}}{\x{}}} = \isprop{\Top}{\makelocal{d}}
to check the body.
With \propenv{} = \isprop{\Top}{\makelocal{d}},
T-If first checks the test \e{1} = {\appexp{\numberhuh{}}{\makelocal{d}}}
via the T-App rule, with three steps.

@@

Resuming {\appexp{\numberhuh{}}{\makelocal{d}}},
in T-App the operand \ep{} = \makelocal{d} is checked with
T-Local as
\begin{equation}
\judgementselfrewrite{\propenv{}}
                     {\makelocal{d}}
                     {\Top}
                     {\filterset{\notprop{\falsy}{\makelocal{d}}}
                                {\isprop{\falsy}{\makelocal{d}}}}
                     {\makelocal{d}}
\end{equation}
which encodes the type, proposition, and object information
about variables.

@@

We now walk through this conversion.% from unresolved expression~\ref{eq:unresolved} to 
%resolved expression~\ref{eq:resolved}.

@@

\begin{figure}
  \footnotesize
  $$
\LUpdate{}
$$
\caption{Update rule}
\label{main:figure:updaterule}
\end{figure}

\begin{figure}
  \footnotesize
$$
\begin{array}{lr}
  \begin{array}{llll}
    \restrictfigure{}
  \end{array}
  \ \ 
  \begin{array}{llll}
    \removefigure{}
  \end{array}
\end{array}
$$
\caption{Restrict and remove}
\label{main:figure:restrictremove}
\end{figure}

@@

A lookup that is not present or absent is checked by
T-GetHMapPartialDefault.
$$
\abs{\makelocal{u}}{\Lunch}{\getexp{\makelocal{u}}{\makekw{bocce}}}
$$
The result type is \Top, and the return object is \path{\keype{\makekw{bocce}}}{\makelocal{u}}.
Notice propositions are erased once they enter a HMap type.

@@

Extending a map with T-AssocHMap preserves its completeness.
$$
\abs{\makelocal{b}}{\Breakfast}{\assocexp{\makelocal{b}}{\makekw{au}}{\makestr{beans}}}
$$
The result type is
$
\HMapc{\mandatoryset{\mandatoryentrynoarrow{\Valkw{en}}{\String}, {\mandatoryentrynoarrow{\Valkw{fr}}{\String}}
        ,{\mandatoryentrynoarrow{\Valkw{au}}{\String}}}}
$,
a complete map.
T-AssocHMap also enforces ${\k{}} \not\in {\absent{}}$ to prevent badly formed types.

@@

\paragraph{Subtyping}
Subtyping for HMaps
designate \MapLiteral{} as a common supertype for all HMaps.
S-HMap says that HMaps are subtypes if they agree
on \completenessmeta{}, agree on mandatory entries with subtyping
and at least cover the absent keys of the supertype.
Complete maps are subtypes of partial maps
as long as they agree on the mandatory entries of the partial map via subtyping (S-HMapP).

### 2015-10-16 overview.tex
shaved 1/4 page

We now begin a tour of the central features of Typed Clojure,
beginning with Clojure itself. In our presentation, we will make 
use of the full Typed Clojure system to illustrate the key type system
ideas, before studying the core features in detail in
\secref{sec:formal}.

@@

Clojure~\cite{Hic08} is a Lisp built to run on the
Java Virtual Machine with exemplary support for concurrent programming
and immutable data structures. It emphasizes mostly-functional
programming, restricting imperative updates to a limited set of
structures which have specific thread synchronization behaviour. By
default, it provides fast implementations of immutable lists, vectors,
and hash tables, which are used for most data structures, although it
also provides means for defining new records.

One of Clojure's primary advantages is easy interoperation with
existing Java libraries. It automatically generates appropriate JVM
bytecode to make Java method and constructor calls, and treats Java
values as any other Clojure value. However, this smooth
interoperability comes at the cost of pervasive \java{null}, which
leads to the possibility of null pointer exceptions---a drawback we
address in Typed Clojure.

@@

The following Typed Clojure program constructs a new \clj{File}
instance and calls the \clj{getParent} method on the result,
returning a string \clj{"a"}, inferred as nullable.

@@

The conditional guards from dereferencing \clj{nil}, and---as before---removing 
it is a static type error, as typed code could possibly dereference \clj{nil}.
\begin{minted}{clj}
(defn parent-bad-in [f :- (U nil File)]
  (.getParent f)) ; Type Error: Cannot call instance method 
                  ; java.io.File/getParent on type (U nil File).
\end{minted}
Since Java-level types do not provide information about nullability,
by default Typed Clojure conservatively assumes that Java method and
constructor arguments, such as \clj{"a/b"} provided to the \clj{File} constructor in
\egref{example:getparent-direct-constructor},
 must be \emph{non-nullable}, but this can be
configured for  particular calls as needed.
The target of method invocations is always non-nullable.

@@

\paragraph{From multimethod to multiple dispatch}
Clojure multimethod dispatch, and Typed Clojure's handling of it, goes
even further, supporting dispatch on multiple arguments via vectors.
%
Dispatch on multiple arguments is beyond the scope of this paper,
but the same intuition applies---adding support for multiple dispatch
automatically allows arbitrary combinations and nestings
of it and previous simple dispatch rules.

>>>>

We now begin a tour of the central features of Typed Clojure,
beginning with Clojure itself. Our presentation
uses the full Typed Clojure system to illustrate key type system
ideas, before studying the core features in detail in
\secref{sec:formal}.

@@

Clojure~\cite{Hic08} is a Lisp that runs on the
Java Virtual Machine with support for concurrent programming
and immutable data structures in a mostly-functional
style.
%, restricting imperative updates to a limited set of
%structures each with specific thread synchronization behaviour. 
%Fast implementations of immutable vectors, and hash tables are featured,
%and a means for defining new records.
%
Clojure provides easy interoperation with
existing Java libraries,
with
Java
values being as any other Clojure value. However, this smooth
interoperability comes at the cost of pervasive \java{null}, which
leads to the possibility of null pointer exceptions---a drawback we
address in Typed Clojure.

@@

The following Typed Clojure program constructs a new \clj{File}
instance and calls the \clj{getParent} method on the result,
returning a nullable string \clj{"a"}.

@@

The conditional guards from dereferencing \clj{nil}, and---as before---removing 
it is a static type error, as typed code could possibly dereference \clj{nil}.
\begin{minted}{clj}
(defn parent-bad-in [f :- (U nil File)]
  (.getParent f)) ; Type Error: Cannot call instance method on nil.
\end{minted}
Since Java-level types do not provide information about nullability,
by default Typed Clojure conservatively assumes that Java method and
constructor arguments, such as \clj{"a/b"} provided to the \clj{File} constructor in
\egref{example:getparent-direct-constructor},
 must be \emph{non-nullable}, but this can be
configured for  particular calls as needed.
The target of method invocations is always non-nullable.

@@

\paragraph{From multimethod to multiple dispatch}
Clojure multimethod dispatch, and Typed Clojure's handling of it, goes
even further, supporting dispatch on multiple arguments via vectors.
%
Dispatch on multiple arguments is beyond the scope of this paper,
but the same intuition applies---adding support for multiple dispatch
admits arbitrary combinations and nestings
of it and previous dispatch rules.

### 2015-10-16 related-work.tex
shaved 1/4 page

% Cite a few of the early papers here.
%http://www.cs.washington.edu/research/projects/cecil/www/pubs/
\paragraph{Multimethods} 
\cite{MS02} and collaborators present a sequence of
systems~\cite{Chambers:1992:OMC,Chambers:1994:TMM,MS02} with statically-typed multimethods
and modular type checking.  In contrast to Typed Clojure, in these
system methods declare the types of arguments that they expect which
corresponds to exclusively using \clj{class} as the dispatch function
in Typed Clojure. However, Typed Clojure does not attempt to rule out
failed dispatches at runtime.

@@

\paragraph{Other optional and gradual type systems}
In addition to Typed Racket, several other gradual type
systems have been developed recently, targeting existing
dynamically-typed languages.  Reticulated Python~\cite{Vitousek14} is
an experimental gradually typed system for Python, implemented as a
source-to-source translation that inserts dynamic checks at language
boundaries and supporting Python's first-class object system.  Typed
Clojure does not support a first-class object system because Java (and
Clojure) have nominal classes, however HMaps offer an alternative to
the structural objects offered by Reticulated. Similarly,
GradualTalk~\cite{gradualtalk} offers gradual typing for SmallTalk,
with nominal classes.

Optional types, requiring less implementation effort and avoiding
runtime cost, have been  adopted in industry, including Hack for PHP~\cite{hack}, and Flow~\cite{flow} and
TypeScript~\cite{typescript}, two extensions of JavaScript. These
systems  support  forms of occurrence typing, but not in the
generality presented here, nor do they include the other features we
present.

>>>>

% Cite a few of the early papers here.
%http://www.cs.washington.edu/research/projects/cecil/www/pubs/
\paragraph{Multimethods} 
\cite{MS02} and collaborators present a sequence of
systems~\cite{Chambers:1992:OMC,Chambers:1994:TMM,MS02} with statically-typed multimethods
and modular type checking.  In contrast to Typed Clojure, in these
system methods declare the types of arguments that they expect which
corresponds to exclusively using \clj{class} as the dispatch function
in Typed Clojure. However, Typed Clojure does not attempt to rule out
failed dispatches.

@@

\paragraph{Other optional and gradual type systems}
%In addition to Typed Racket, 
Several other gradual type
systems have been developed for existing
dynamically-typed languages.  Reticulated Python~\cite{Vitousek14} is
an experimental gradually typed system for Python, implemented as a
source-to-source translation that inserts dynamic checks at language
boundaries and supporting Python's first-class object system. 
Clojure's nominal classes avoids the need to support
first-class object system in Typed Clojure, however HMaps offer an alternative to
the structural objects offered by Reticulated. Similarly,
GradualTalk~\cite{gradualtalk} offers gradual typing for SmallTalk,
with nominal classes.

Optional types
%, requiring less implementation effort and avoiding
%runtime cost, 
have been  adopted in industry, including Hack for PHP~\cite{hack}, and Flow~\cite{flow} and
TypeScript~\cite{typescript}, two extensions of JavaScript. These
systems  support  limited forms of occurrence typing,
and do not include the other features we
present.
